{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}

### 2018-01-03 thesis/thesis.tex
write some inspirational quotes

% This is a template for a CMU thesis.  It is 18 pages without any content :-)
% The source for this is pulled from a variety of sources and people.
% Here's a partial list of people who may or may have not contributed:
%
%        bnoble   = Brian Noble
%        caruana  = Rich Caruana
%        colohan  = Chris Colohan
%        jab      = Justin Boyan
%        josullvn = Joseph O'Sullivan
%        jrs      = Jonathan Shewchuk
%        kosak    = Corey Kosak
%        mjz      = Matt Zekauskas (mattz@cs)
%        pdinda   = Peter Dinda
%        pfr      = Patrick Riley
%        dkoes = David Koes (me)

@@

% some useful packages
\usepackage{times}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[charter]{mathdesign}
\usepackage[numbers,sort]{natbib}
\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
%pdfborder=0 0 0,  %removes outlines around hyper links in online display
]{hyperref}
\usepackage{subfigure}
\usepackage[linesnumbered,ruled]{algorithm2e}
\usepackage{algpseudocode}

@@

\newcommand\simics{\textsc{Simics}}
\newcommand{\sect}[1]{\S #1}
\newcommand\hilight[2]{\color{#1}#2\color{black}}
\definecolor{orange}{RGB}{192,96,0}
\definecolor{olivegreen}{RGB}{0,127,0}
\definecolor{brickred}{RGB}{192,0,0}
\definecolor{commentblue}{RGB}{0,0,192}

\newtheorem{theorem}{Theorem}

\newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
\newcommand\llitem{\item[\llama{1.2em}]}

@@

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Practical Concurrency Testing} \\
\normalsize \vspace{1em}
{\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}}
\author{Ben Blum}
\date{January 2017}
\Year{2017}
\trnumber{}

\committee{
Garth Gibson, Chair \\
David A. Eckhardt \\
Brandon Lucia \\
Haryadi Gunawi (University of Chicago)
}

@@

% TODO(thesis): Save this one for the thesis itself...
%\keywords{landslide terminal, baggage claim, ground transportation, ticketing}

@@

%\begin{dedication}
%	For my dog, Louie.
%\end{dedication}

@@

In this thesis, I propose to make stateless model checking more practical for human use by way of several new techniques.
I have built Landslide, a stateless model checker specializing in student projects for undergraduate operating systems classes.
Landslide includes a novel algorithm for automatically managing
%the exploration of
multiple state spaces according to their estimated completion times,
which I will show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
I will evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class.
%Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
Finally, I will explore broader impact by extending Landslide to test some real-world programs and to be used by students at other universities.
\end{abstract}

%\begin{acknowledgments}
%My advisor is cool.
%
%But I am cooler.
%\end{acknowledgments}

\tableofcontents
%\listoffigures
%\listoftables

@@

@@

Modern computer architectures have turned to increasing CPU core count, rather than clock speed, to improve processing power \cite{mooreslaw}.
%To take advantage of multiple cores, programs must be written {\em concurrently}
To take advantage of multiple cores for performance, programmers must write software to execute {\em concurrently} --
using multiple {\em threads} which execute multiple parts of a program's logic simultaneously.
However, when threads access the same shared data, they may interleave in unexpected ways which change the outcome of their execution.
When an unexpected interleaving produces undesirable program behaviour,
for example, by corrupting shared data structures,
we call it a {\em concurrency bug}.
Concurrency bugs are notoriously hard for programmers to find and debug
because the specific thread interleaving required to trigger them arises at random during normal execution,
and often with very low probability.
%Concurrency bugs are notoriously hard to find and reproduce because they only appear in specific thread interleavings, which arise at random during normal program execution.
% TODO(LAYMAN): give example of trying to open car door at same time as friend turns key to unlock it.

Most commonly, a programmer searches for concurrency bugs in her code by running it many times (in parallel, in serial, or both),
hoping that eventually, it will run according to the particular interleaving required to expose a hypothetical bug.
This technique, known as {\em stress testing}, is unreliable,
providing no guarantee of finding the failing interleaving in any finite amount of time.
%, even when a bug does exist.
It also provides no assurance of correctness:
when finished, there is no way of knowing how many distinct thread interleavings were actually tested.
%it may by chance test only a single interleaving over and over again!
Nevertheless, stress testing remains popular because of how easily a programmer can use it:
she simply wraps her program in a loop, sets it to run overnight, and kills it if her patience runs out before it finds a bug.

{\em Stateless model checking} \cite{verisoft} is an alternative way to test for concurrency bugs,
or to verify their absence,
which provides more reliable coverage, progress, and verification than stress testing.
A stateless model checker tests a program by forcing it to execute a new unique thread interleaving on each iteration of the test,
capturing and controlling the randomness in a finite state space of all possible interleavings.

Unfortunately, the size of these state spaces is exponentially proportional to the size of the tested program.
% TODO(LAYMAN): explain exponential explosion by relating the parable of grains of rice on a chessboard.
For even moderately-sized programs, there may be more possible ways to interleave every thread's every instruction
than particles in the universe.
Accordingly, a programmer who wants her test to make reasonable progress through the state space must choose a subset of ways that her threads could interleave,
focusing on fully testing that subset, while ignoring other possibilities she doesn't care about.
%However, choosing this subset is a difficult trade-off for humans to make before even knowing whether a bug exists.
However, it is difficult to choose a subset of thread interleavings that will produce a meaningful, yet feasible test.
Until computers can automatically navigate this trade-off in some intelligent way,
programmers will continue to fall back to the random approach of stress testing.

Another problem stateless model checking suffers is that certain types of programs cannot be tested without the programmer putting forth some manual instrumentation effort.
For example, operating system kernels implement their own sources of concurrency and their own synchronization primitives,
so the checker needs to be told how to identify and control the execution of each thread.
Some expert concurrency research wizards may be willing to add manual annotations to their code,
but required manual effort is a serious downside for anyone with a looming deadline,
and especially so for students who are still learning basic concurrency principles.
%We should not expect programmers to add effortful manual annotations to their code,
%or they will abandon our fancy technique to instead simply run stress tests until their deadline tomorrow evening.

This thesis will solve both problems discussed above.
My thesis statement is as follows:

\vspace{1em}

\begin{center}
	{\em Thanks to the new algorithms, heuristics, and concurrency models I have developed,
	stateless model checking is an appropriate and accessible concurrency testing technique
	for programmers in both educational and real-world settings.}
\end{center}

\vspace{1em}

I have built Landslide \cite{landslide}, a stateless model checker for thread libraries and kernels,
and I have developed some techniques for automatically choosing the best thread interleavings to test
and for automatically instrumenting operating system kernels in an educational setting.
This thesis will comprise three major contributions:

\begin{itemize}
	\item {\bf Meaningful state spaces (Chapter \ref{chap:quicksand}).}
		I will present {\em Iterative Deepening}, a new algorithm for navigating the trade-off in how many preemption points to test at once.
		Iterative Deepening incorporates state space estimation \cite{estimation} to decide on-the-fly whether each state space is worth pursuing, and uses data race analysis \cite{tsan} to find new preemption point candidates based on a program's dynamic behaviour.
		This section will include a large evaluation of the technique, comparing its performance to three prior work approaches across 600+ unique tests.
		I will show that Iterative Deepening of preemption points outperforms prior work in terms both of finding bugs quickly and of completely verifying correctness when no bug exists.

		This work is {\bf completed} and was published at OOPSLA 2016.
	\item {\bf Educational use (Chapters \ref{chap:410} and \ref{chap:pintos}).}
		For the past three semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.
		I will continue these user studies, and use the data to evaluate the suitability of stateless model checking in an educational setting.
		%, investigating the following questions in particular:
		%\begin{itemize}
		%	\item Does Landslide improve project submission quality when used as a debugging aid?
		%	\item Does Landslide teach students lessons in concurrent programming
		%\end{itemize}

		So far, the fully-automatic testing mode is available only for 15-410 thread library projects.
		To prove these techniques are relevant beyond CMU's walls, I will extend Landslide to handle Pintos kernel projects from other universities \cite{pintos}.
		I will then collaborate with those schools to deploy Landslide to their operating systems students.

		This is {\bf ongoing work}; I have run the user study for 4 semesters so far and am proposing to continue them
		and extend the existing evaluation,
		as well as deploy a Pintos port.
	\item {\bf Transactional Memory (Chapter \ref{chap:tm}).}
		Transactional Memory (TM) is a relatively new concurrent programming technique \cite{transactional-memory}
		which is not yet addressed by modern model checkers.
		I will extend Landslide's concurrency model to support both hardware (HTM) and software (STM) variants of TM,
		and test several ``real-world'' TM programs.

		This will be {\bf new work},
		involving some implementation effort, some proofs, and designing a new evaluation.
		%, as Landslide presently supports testing Pintoses only with some manual effort,
		%and does not yet feature instrumentation to run on Linux programs.
		%and will need its concurrency model extended to incorporate the HTM programming model.
\end{itemize}

% TODO(thesis): Other chapters of this thesis include...

\input{overview}
\input{quicksand}
\input{410}
\input{pintos}
\input{tm}
\input{related}

%Reduction
%techniques such as Dynamic Partial Order Reduction \cite{dpor} and Maximal Causality Reduction \cite{mcr} expand the limits of feasible test completion,
%and search ordering strategies such as Iterative Context Bounding \cite{chess-icb} \revision{encourage}~bugs to be found sooner in a given space should they exist.

%Topics of thesis:
%1. Data race preemption points / iterative deepening
%2. Education in 410. Evaluation questions are:
%	2a. Does P2 submission quality improve?
%	2b. Does P3 submission quality improve correlated with students who used (got the most out of) Landslide during P2?
%	2c. Does Landslide reach the bottom of the class? (Look at bottom-of-class P0s/P1s, see how they improve)
%3. Broader impact
%	3a. Real world programs (and the challenges therein?)
%	3b. Give to pintos kids (and open-source landslide on bochs in doing so)

\chapter{Conclusion and Timeline}

Stateless model checking is a powerful technique for testing concurrent programs,
capable in theory of rooting out any nondeterministic bug or providing total verification on any program,
but suffers several problems which relegate that theory to fantasy land \cite{vargomax}.
Chief among those problems is exponential explosion of state spaces,
making it difficult to decide in advance which combinations of preemption points can be productively tested within a given time limit.
Another major problem is the manual annotation effort required to test certain types of concurrent programs,
which is especially relevant for operating systems students who implement their own kernels.
In this document I have proposed a thesis which will solve both problems.
I leave you now with a proposed timeline for bringing each project to fruition.

\begin{itemize}
	\item 2017 January: Submit thesis proposal, submit revised IRB proposal for planned user studies.
	\item 2017 ???uary: Present thesis proposal (whenever the committee is free).
	\item 2017 February-March: Deploy 15-410 student user study for more data (Section~\ref{sec:studence}).
	\item 2017 March-August: Port Landslide to Bochs for use in Pintos OS classes (Section~\ref{chap:pintos}).
	\item 2017 March-August: Implement, evaluate, and write paper about transactional memory project (Section~\ref{chap:tm}).
	\item 2017 August-October: Polish and deploy Pintos user study (Section~\ref{chap:pintos}).
	\item 2017 ???ember: Resubmit paper about 410 user studies (Section~\ref{sec:studence}) (previously rejected from SIGCSE) (whenever an appropriate conference deadline arises).
	\item 2017 October-November: Deploy 410 student study again for more data.
	\item 2017 December: Begin writing thesis.
	\item 2018 January-April: Write thesis. Also fill in miscellaneous gaps, such as the ``variance'' Quicksand experiments (Section~\ref{chap:quicksand}) and measuring Landslide's heuristics' accuracy (Section~\ref{sec:studence}).
	\item 2018 February-March: Deploy Pintos student and 410 TA studies one final time.
	\item 2018 April-May: Defend \& graduate.
\end{itemize}

%For pretty much any reason (much like hardware transactions),
%I may need to delay the timeline by a semester, defending instead in 2018 June.
%For example, the above plan to deploy a Pintos user study within one semester's time of finishing the open-source Landslide port is quite ambitious,
%assuming among other things smooth negotiations with the OS course staff at another university.
%Anyway, this is my optimistic proposal. Thanks for reading.

%\appendix
%\include{appendix}

@@

% By default \bibsection is \chapter*, but we really want this to show
% up in the table of contents and pdf bookmarks.
\renewcommand{\bibsection}{\chapter{\bibname}}
%\newcommand{\bibpreamble}{This text goes between the ``Bibliography''
%  header and the actual list of references}
\bibliographystyle{plainnat}
\bibliography{citations} %your bib file

>>>>

% This is a template for a CMU thesis.  It is 18 pages without any content :-)
% The source for this is pulled from a variety of sources and people.
% Here's a partial list of people who may or may have not contributed:
%
%        bnoble   = Brian Noble
%        caruana  = Rich Caruana
%        colohan  = Chris Colohan
%        comar    = Cyrus Omar
%        jab      = Justin Boyan
%        josullvn = Joseph O'Sullivan
%        jrs      = Jonathan Shewchuk
%        kosak    = Corey Kosak
%        mjz      = Matt Zekauskas (mattz@cs)
%        pdinda   = Peter Dinda
%        pfr      = Patrick Riley
%        dkoes = David Koes (me)

@@

% some useful packages
\usepackage[charter]{mathdesign}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[numbers,sort]{natbib}
\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
%pdfborder=0 0 0,  %removes outlines around hyper links in online display
]{hyperref}
\usepackage{subfigure}

@@

@@

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Practical Concurrency Testing} \\
\normalsize \vspace{1em}
{\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}}
\author{Ben Blum}
\date{Decembruary 2018}
\Year{2018}
\trnumber{}

\committee{
Garth Gibson, Chair \\
David A. Eckhardt \\
Brandon Lucia \\
Haryadi Gunawi, University of Chicago
}

@@

\keywords{landslide terminal, baggage claim, ground transportation, ticketing}

@@

\begin{dedication}
For Elsie, Margie, Johnny, Mel, Eve, and David, who brought me here today.
\end{dedication}

@@

In this thesis, I develop several new techniques to make stateless model checking more practical for human use.
I have built Landslide, a stateless model checker specializing in undergraduate operating systems class projects.
Landslide extends the traditional model checking algorithm with a new framework for automatically managing
%the exploration of
multiple state spaces according to their estimated completion times,
which I show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
I evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class,
and more recently, students in the operating systems classes at the University of Chicago and Berkeley.
%Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
%Finally, I will explore broader impact by extending Landslide to test some real-world programs and to be used by students at other universities.
Finally, I extend Landslide with a new concurrency model for hardware transactional memory,
and evaluate several real-world transactional benchmarks to show that stateless model checking can keep up with the developing concurrency demands of real-world programs.
\end{abstract}

\begin{acknowledgments}
I thank the online Android: Netrunner community Stimhack.
\\

Please hold on. This thesis will be departing shortly for the Landslide terminal, baggage claim, ground transportation, and ticketing.
\end{acknowledgments}

\newpage
\begin{figure}[h]
	\vspace{12em}
	\begin{center}
		\includegraphics[width=0.5\textwidth]{how-i-learned.png}
	\end{center}
\end{figure}

\tableofcontents
\listoffigures
\listoftables

@@

\newcommand{\inspirationallinebreak}{\vspace{0.25em}}
\newcommand\inspirationalquote[2]{\begin{flushright}
	\vspace{-1em}
	\fontfamily{pzc}\selectfont
	{\em #1}
	\inspirationallinebreak

	---{#2}
	\fontfamily{mdbch}\selectfont
	\vspace{2em}
\end{flushright}}

@@

\chapter{Background}

\chapter{410} % rename
\inspirationalquote{Knowing the students might one day
%find a way to
fix their concurrency bugs...
it fills you with determination.}{Undertale (paraphrased)}

\chapter{Related Work}
\inspirationalquote{I see now that none of us are yet ready. The cycle exists so that we may
improve ourselves. But \\
the one who reaches the summit is not our superior, for
they stand on our shoulders to reach it.}
{The Shepherd v82.6.0174, The Talos Principle}

\chapter{Conclusion}
\inspirationalquote{
%	``If you are a god, Zeus, as the stories claim, then why did you create evolution?
%	Why did you make \\ a world that can only grow through cruelty and pain?''
%	\inspirationallinebreak
%
%For a moment--surely this meant death was near--she thought she heard him answer:
%``My child, I \\
%was shaped by the gods that came before me, as you were shaped by me.
%The choice I had was be- \\
%tween creation and oblivion, life and death.
%And I chose life, because any life is better than no life, \\
%because as long as there is life, there is hope--if not for us, then for some generation to come.''
%	\inspirationallinebreak
%
%``Then how are you a god, if you can offer so little?'' she whispered, feeling death creep closer.
%	\inspirationallinebreak
%
%``I am a god because I take upon myself the burden of creation,'' the statue replied.
%	\inspirationallinebreak
%
%``Then we are all gods,'' Alexandra said, and pushed the button.}
	\begin{tabular}{p{0.875\textwidth}}
	``If you are a god, Zeus, as the stories claim, then why did you create evolution?
	Why did you make a world that can only grow through cruelty and pain?''
	\\
	For a moment--surely this meant death was near--she thought she heard him answer:
	``My child, I was shaped by the gods that came before me, as you were shaped by me.
	The choice I had was between creation and oblivion, life and death.
	And I chose life, because any life is better than no life,
	because as long as there is life, there is hope--if not for us, then for some generation to come.''
	\\
	``Then how are you a god, if you can offer so little?'' she whispered, feeling death creep closer.
	\\
	``I am a god because I take upon myself the burden of creation,'' the statue replied.
	\\
	``Then we are all gods,'' Alexandra said, and pushed the button.
	\end{tabular}}
{The Talos Principle (Road to Gehenna)}

asdfasdasdf
%\appendix
%\include{appendix}

@@

% By default \bibsection is \chapter*, but we really want this to show
% up in the table of contents and pdf bookmarks.
\renewcommand{\bibsection}{\chapter{\bibname}
\inspirationalquote{Only a doctor of philosophy, Darth.}{Robert Marsh, via Twitter}
}
%\renewcommand{\bibpreamble}{This text goes between the ``Bibliography''
%  header and the actual list of references}
\bibliographystyle{plainnat} % plain n'at
\bibliography{register} %your bib file
